---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "svg"
)
```
# cannonball

cannonball bundles a couple of functions that I use
when teaching introductory courses 
in quantitative methodology and statistics.

## Installation

You can install cannonball from Github with:

``` r
pak::pak("janhove/cannonball")
```

## Functions

```{r}
# Load the package
library(cannonball)
```

### `plot_r()`: Draw different scatterplots with the same correlation coefficient

A single correlation coefficient can correspond to any number of scatterplots.
`plot_r()` produces 16 such scatterplots for any given Pearson correlation 
coefficient to drive this point home.

```{r example}
plot_r(r = 0.5, n = 35)
```

For more info, see `?plot_r`.

Accompanying blog post: [_What data patterns can lie behind a correlation coefficient?_](https://janhove.github.io/posts/2016-11-21-what-correlations-look-like/)

### Walkthroughs

To help students see the connection between an experiment's
design and its analysis, I've written two functions.

`walkthrough_p()` guides the user through a completely randomised
experiment: Data points are generated and randomly assigned
to either the control or intervention condition. Then, the
intervention effect is added to the data points in the intervention
condition. Finally, the data are analysed using a randomisation test.

```{r, eval = FALSE}
# see ?walkthrough_p
walkthrough_p(n = 18, diff = 0.3, sd = 1,)
```

`walkthrough_blocking()` works similarly to `walkthrough_p()`
but describes a randomised block design: Prior information about
the data points is available in the form of a covariate (e.g.,
a pretest score). This information is used to group participants
into 'blocks', and the randomisation is restricted in that one
participant per block is assigned to the control and one to the
intervention condition. Crucially, the analysis needs to take this
restricted randomisation into account.

```{r, eval = FALSE}
# ?walkthrough_blocking
walkthrough_blocking(n = 12, diff = 0.4, sd = 1)
```

### Simulate data and analysed cluster-randomised data
The data from experiments in which entire clusters
of participants (e.g., classes) are assigned to the
experimental conditions can't be analysed in the same
way as data from experiments in which the participants
are assigned to the conditions individually. `clustered_data()`
generates data for a cluster-randomised experiment and can
be used to demonstrate the increased Type-I error rate
if such data are analysed using t-tests on the individual outcomes.

Refer to the vignette (article) `vignette("cluster-randomisation", package = "cannonball")`
for details.

### Check model assumptions graphically
These functions may be helpful for helping you to judge
whether your data conform to the assumptions of your statistical model.
By embedding the model's diagnostic plot in a line-up of diagnostic plots
of simulated data for which the model's assumptions are literally met,
analysts can more easily determine whether any blips in these plots are
indicative of assumption violations or whether they can plausibly be
accounted for by sampling error/noise. 

Refer to the vignette (article) `vignette("check-assumptions", package = "cannonball")`
for details.
