---
title: "Graphically checking model assumptions"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_align: "center"
    dev: svg
vignette: >
  %\VignetteIndexEntry{check-assumptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `cannonball` package contains a handful of functinos that can help
you judge whether your data conform to the assumptions of the statistical
model you've fitted.
By embedding the model's diagnostic plot in a line-up of diagnostic plots
of simulated data for which the model's assumptions are literally met,
you can more easily determine whether any blips in these plots are 
indicative of assumption violations or whether they can
plausibly be accounted for by sampling error or noise.
The idea for this stems from [Buja et al. (2009)](http://doi.org/10.1098/rsta.2009.0120);
see [Vanhove (2018)](https://doi.org/10.31234/osf.io/zvawb) for an accompanying article.

## Example with numeric predictors
Load the package and fit a simple regression model.

```{r setup}
library(cannonball)

m <- lm(mpg ~ wt, data = mtcars)
```

Using `parade()`, create a parade in which the real dataset is hidden
between 19 other datasets generated from the model.
For these other datasets, the model's assumptions are literally met.

```{r}
my_parade <- parade(m)
```

The parade is itself a tibble/data frame:

```{r}
my_parade
```

A handful of convenience functions are available for gauging how much
the true dataset stands out from the simulated ones.
For instance, `lin_plot()` can be used to check if the linearity assumption
is met: Which of the plots below looks most different from the rest?

```{r}
lin_plot(my_parade)
```
In nineteen of the above plots, the linearity assumption is literally met
yet the scatterplot smoothers are all nonlinear to some extent.
If the true relationship were linear, there should only be a one-in-twenty
chance that the plot you picked is the one with the true data.

Using `reveal()`, you can check your guess:
```{r}
reveal(my_parade)
```

You can also plot the residuals against a predictor like so:

```{r}
lin_plot(my_parade, "wt")
```


Related functions are:

* `var_plot()`: Check for non-constant variance in the residuals.
* `norm_qq()`: Check for non-normality of the residuals using a quantile--quantile plot.
* `norm_hist()`: Check for non-normality of the residuals using a histogram.

```{r}
my_parade <- parade(m)
var_plot(my_parade)
# reveal(my_parade) # check

my_parade <- parade(m)
norm_qq(my_parade)
# reveal(my_parade)

my_parade <- parade(m)
norm_hist(my_parade)
# reveal(my_parade)
```

## Example with categorical predictors
If you're working with categorical predictors, the standard `var_plot()`
is a bit difficult to read.
One option is to draw, say, boxplots per cell:

```{r}
mtcars$f.gear <- factor(mtcars$gear)
m <- lm(mpg ~ f.gear, data = mtcars)
my_parade <- parade(m)


library(ggplot2)
my_parade |> 
  ggplot(aes(x = f.gear, y = .resid)) +
  geom_boxplot() +
  facet_wrap(vars(.sample))

# or even
my_parade |> 
  ggplot(aes(x = f.gear, y = .resid)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(shape = 1, position = position_jitter(width = 0.2, height = 0)) +
  facet_wrap(vars(.sample))
```

Alternatively, or additionally, `parade_summary()` can be used to compute
summary statistics of the residuals per cell; `var_plot()` can then plot these.
This is particularly useful if the outcome variable is pretty coarse:
Without the by-cell averaging, you would be able to identify the true data
based not on violations of the homoskedasticity assumptions but based on
the fairly coarse nature of the data.

```{r}
my_parade |> 
  parade_summary() |> 
  var_plot()
```

## References
Buja, Andreas, Dianne Cook, Heike Hofmann, Michael Lawrence, Eun-Kyung Lee, Deborah F. Swayne and Hadley Wickham.
2009.
[Statistical inference for exploratory data analysis and model diagnostics.](http://doi.org/10.1098/rsta.2009.0120)
_Philosophical Transactions of the Royal Society A_ 367(1906).
4361--4383.

Vanhove, Jan.
2018.
[Checking the assumptions of your statistical model without getting paranoid](https://doi.org/10.31234/osf.io/zvawb).
Preprint on PsyArxiv.
