---
title: "Simulating and analysing cluster-randomised designs"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_align: "center"
    dev: svg
vignette: >
  %\VignetteIndexEntry{cluster-randomisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The data from experiments in which entire clusters of participants 
(e.g., classes) are assigned to the experimental conditions 
can’t be analysed in the same way as data from experiments 
in which the participants are assigned to the conditions individually. 
The function `clustered_data()` generates data for a 
cluster-randomised experiment and can be used to demonstrate the 
increased Type-I error rate if such data are analysed using t-tests 
on the individual outcomes.
See [Vanhove (2015)](http://doi.org/10.14746/ssllt.2015.5.1.7)
for an introduction geared towards applied linguists.

## Simulating cluster-randomised data
First load the package.
```{r setup}
library(cannonball)
```

The `n_per_class` parameter specifies the number of pupils in each simulated
class. Half of the classes are assigned to the intervention condition and half 
to the control condition.
```{r}
d <- clustered_data(
  n_per_class <- c(17, 26, 14, 18, 19, 22, 17, 21),
  ICC = 0.25,
  effect = 0.3
)
d
xtabs(~ class + condition, d)
```

I like to plot the outcomes of cluster-randomised experiments with a continuous
outcome like so:

```{r}
library(ggplot2)
ggplot(data = d,
       aes(x = reorder(class, outcome, FUN = median),
           y = outcome)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(shape = 1,
             position = position_jitter(width = 0.2)) +
  facet_wrap(~ condition, scales = "free_x") +
  xlab("Class") +
  ylab("Outcome")
```

Or like so:

```{r}
ggplot(data = d,
       aes(x = reorder(class, outcome, FUN = median),
           y = outcome)) +
  geom_boxplot(
    outlier.shape = NA,
    mapping = aes(fill = condition)
  ) +
  geom_point(shape = 1,
             position = position_jitter(width = 0.2)) +
  xlab("Class") +
  ylab("Outcome") +
  theme(legend.position = "bottom")
```

If you want to generate data for a cluster-randomised design that includes
a pretest, you can specify the desired population-level correlation
between the pretest scores and the non-intervention posttest scores
via the `rho_prepost` parameter:

```{r}
d <- clustered_data(
  n_per_class <- c(17, 26, 14, 18, 19, 22, 17, 21),
  ICC = 0.25,
  effect = 0.3,
  rho_prepost = 0.8
)
d
```

```{r}
ggplot(
  d,
  aes(x = pretest, y = outcome,
      colour = condition) 
) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

See [Vanhove (2020)](https://doi.org/10.31234/osf.io/ef4zc) for more information
on leveraging pretest scores (and covariates in general) when designing
and analysing data from cluster-randomised designs.

## Evaluating the Type-I error rate of naive analysis
To drive the point home that the data from clustered-randomised designs
can't ignore the clustered nature of the data, we can run a simulation
that generates a couple of thousand such data sets and analyses them
using a t-test on the pupils:
```{r}
class_n <- c(17, 26, 14, 18, 19, 22, 17, 21)
ps <- replicate(
  5000,
  {
    d <- clustered_data(n_per_class = class_n, ICC = 0.25, effect = 0)
    t.test(outcome ~ condition, data = d, var.equal = TRUE)$p.value
  }
)
```

The Type-I error rate for these settings is over 40% as opposed to the 
nominal 5% (when using $\alpha = 0.05$):
```{r}
data.frame(p = ps) |> 
  ggplot(aes(x = ps)) +
  geom_histogram(breaks = seq(0, 1, 0.05),
                 colour = "black",
                 mapping = aes(fill = ps < 0.05)) +
  scale_fill_manual(values = c("#2b83ba", "#d7191c")) +
  xlab("p-value") +
  ylab("Frequency") +
  ggtitle("Inflated Type-I error rate") +
  theme(legend.position = "none")
```

## A correct analysis
One possible way to analyse cluster-randomised data correctly is to
average the outcomes per cluster and then just analyse these cluster averages.

```{r}
d <- clustered_data(
  n_per_class <- c(17, 26, 14, 18, 19, 22, 17, 21),
  ICC = 0.25,
  effect = 0.3
)
d_per_class <- aggregate(d$outcome, 
                         by = list(condition = d$condition, class = d$class), 
                         mean)
d_per_class
```

```{r}
t.test(x ~ condition, d_per_class, var.equal = TRUE)
```

We can verify that this analysis is valid using a simulation:

```{r}
ps <- replicate(
  5000,
  {
    d <- clustered_data(n_per_class = class_n, ICC = 0.25, effect = 0)
    d_per_class <- aggregate(d$outcome, 
      by = list(condition = d$condition, class = d$class), mean)
    t.test(x ~ condition, data = d_per_class, var.equal = TRUE)$p.value
  }
)
data.frame(p = ps) |> 
  ggplot(aes(x = ps)) +
  geom_histogram(breaks = seq(0, 1, 0.05),
                 colour = "black",
                 mapping = aes(fill = ps < 0.05)) +
  scale_fill_manual(values = c("#2b83ba", "#d7191c")) +
  xlab("p-value") +
  ylab("Frequency") +
  ggtitle("No Type-I error rate inflation") +
  theme(legend.position = "none")
```

## References
Vanhove, Jan. 
2015. 
[Analyzing randomized controlled interventions: Three notes for applied linguists.](http://doi.org/10.14746/ssllt.2015.5.1.7)
_Studies in Second Language Learning and Teaching_ 5(1). 
135–152. 
(Also see the [correction note](http://pressto.amu.edu.pl/index.php/ssllt/article/view/5827/5895) for this article.)

Vanhove, Jan.
2020.
[Capitalising on covariates in cluster-randomised experiments.](https://doi.org/10.31234/osf.io/ef4zc)
_PsyArXiv Preprints_.
